### 2023-02-28 calculus.tex
Move marking of numbers to front of formal development

The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
type error localization in bidirectional systems, a judgmental framework provides a unified theory
for the design of typing semantics that supports ill-typed programs. \Cref{sec:calculus-calculus}
describes its form and metatheory by example on a minimal gradually typed lambda calculus, extending
with binary product types in \cref{sec:calculus-products}. All rules and theorems may be found in
the supplementary appendix, alongside a complete mechanization in the Agda proof assistant, which is
discussed in \cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the
system might be extended to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, as
shown in \cref{fig:calculus-syntax-uexp}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expression. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
is the unknown type from gradual type theory.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \textsf{Type} & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \textsf{Exp}  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \\
         &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
                   \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
  \end{array}\]
  \caption{Syntax of the extended gradually typed lambda calculus}
  \label{fig:calculus-syntax-uexp}
\end{figure}

The language is typed bidirectionally, which employs two mutually defined judgments. \emph{Type
synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context
$\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected. We may give these typing semantics traditionally, without considering error
localization and recovery.

The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are \emph{unmarked expressions}.

@@

  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
    inserting error markings where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery.

@@

We give, for now, only part of the marked expression syntax in \cref{fig:calculus-syntax-mexp-part}.
Throughout the remainder of this section, as we consider the formulation of marking, we will add the
necessary error mark forms.

% TODO include holes
\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \cdots
             % &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Partial syntax of marked expressions}
  \label{fig:calculus-syntax-mexp-part}
\end{figure}

@@

How can we ensure that the marking procedure is correctly defined? There are two critical
metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}
%
That is, we may mark, i.e. type check and localize errors for \emph{any} syntactically well-formed
program. Furthermore, marking should preserve the syntactic structure of programs, modulo error
marks. Thus, we must define a notion of \emph{mark erasure}, which converts marked expressions back
into unmarked expression by removing error marks. \Cref{fig:calculus-mark-erasure-part} gives the
definition of mark erasure, written $\erase{\ECMV}$, for our initial marked expression syntax. Then,
we may define a \emph{well-formedness} criterion for marking:
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
marked expressions that are well-typed; indeed, their types should match the types synthesised or
analysed against in marking. % TODO more intuition for this theorem

\input{figs/calculus-mark-erasure-part}

@@

@@

% TODO subsumption is discussed later...
The typing of unmarked variables in the gradually typed lambda calculus is standard. Because of
subsumption, we need only define a synthesis rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking would convert the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{ISVar}

@@

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound? Then, \textsc{USVar} does not apply, and the typing semantics say nothing about such
expressions. To satisfy totality, such a possibility motivates an \emph{unbound variable mark}:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECUnbound{x}
\end{array}\]
%
accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{ISUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing.

To prove \cref{thm:calculus-marking-well-formedness}, we extend the definition of mark erasure to
handle the new mark:
%
\[\begin{array}{rcl}
  & \cdots & \\
  \erasesToRow{\ECUnbound{x}}{x}
\end{array}\]

@@

There are explicit synthesis and analysis rules for lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark}, which indicates that the expression does not
synthesize a type consistent with the expected type:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconType{\ECMV}
\end{array}\]
%
along with the following rules, where the lambda abstraction is placed inside the inconsistent type
mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2*}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analysed against the unknown type in \textsc{IALam2}.

@@

Finally, as with the unbound variable mark, we extend mark erasure:
%
\[\begin{array}{rcl}
  & \cdots & \\
  \erasesToRow{\ECInconType{\ECMV}}{\erase{\ECMV}}
\end{array}\]

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as binary products. Indeed, in \cref{sec:calculus-products}, we opt for this
approach. Another approach is to mark an error around the entire application expression.

The marking and typing of numbers is straightforward:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynType{\ctx}{\ENumMV}{\TNum}
  }{USNum}

  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{ISNum}

For addition expressions, we simply recursively mark each operand, which must analyse against
$\TNum$, and build the marked addition expression. In this way, blame for type errors is shift into
the operands:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\\\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}

  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{ISPlus}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}
%
Again, we need only synthetic rules; subsumption handles the analytic case.

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
the marked if-else expression from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{ISIf}

@@

However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}:
%
\[\begin{array}{rrcl}
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
\end{array}\]

Then, adding the following rules ensures marking totality on conditionals:
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{ISInconsistentBranches}

@@

The extension of mark erasure is again straightforward:
%
\[\begin{array}{rcl}
  & \cdots & \\
  \erasesToRow{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\EIf{(\erase{\ECMV_1})}{(\erase{\ECMV_2})}{(\erase{\ECMV_3})}}
\end{array}\]

@@

\subsection{Reified Hazelnut Action Semantics}

>>>>

The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
type error localization in bidirectional systems, a judgmental framework that provides a unified
theory for the design of typing semantics that supports ill-typed programs.
\Cref{sec:calculus-calculus} describes its form and metatheory by example on a minimal gradually
typed lambda calculus. All rules and theorems may be found in the supplementary appendix, alongside
a complete mechanization in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the system might be
extended to more complex typing features, such as System F-style polymorphism.

@@

We now formally introduce the \emph{marking calculus} through its application to a gradually typed
lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
$\TUnknown$ is the unknown type from gradual type theory.

% TODO include holes
\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV} \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of the extended gradually typed lambda calculus}
  \label{fig:calculus-syntax}
\end{figure}

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

@@

  \item \textbf{Marking}, which \emph{marks}, i.e. type checks and localizes errors for, any unmarked
    expression into a marked expression, inserting error marks where appropriate.
\end{itemize}
%
In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services. % such as THI.

@@

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally three kinds of
\emph{error marks}, which represent the different type errors that may arise.
%
\begin{itemize}
  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    $x$, as in \cref{fig:calculus-examples-unbound}.

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    expression does not match an expected type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    that the branches of a conditional have inconsistent types.
\end{itemize}
%
Throughout the remainder of this section, as we consider the formulation of marking, we will
motivate and give precise semantics for these error marks. Note that this is a non-exhaustive list;
depending on the syntactic forms and typing features found in the language, other kinds of marks may
be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
required in an language with explicit polymorphism.

@@

How can we ensure that the marking procedure is correctly defined? There are two critical
metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}
%
That is, we may mark \emph{any} syntactically well-formed program.

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
well-typed marked expressions; indeed, their types should match the types synthesised or analysed
against in marking. % TODO more intuition for this theorem

\input{figs/calculus-mark-erasure}

@@

% TODO need to say something about subsumption and how it will be handled later.
To start, let us consider the simple case of numbers. Because of subsumption, we need only define a
synthesis rule for unmarked numbers, in which numbers synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynType{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
only need the following rules for marking unmarked numbers and typing marked numbers:
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{ISNum}

In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
thus analyse the operands against $\TNum$, giving the following rule:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{ISPlus}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

@@

Let us now consider the more interesting case of variables. The typing of unmarked variables in the
gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
synthetic rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{ISVar}

@@

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{ISUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing---and the unbound variable to be used in any
expression. % TODO clarify why this is correct

@@

There are explicit synthesis and analysis rules for unmarked lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{IALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analysed against the unknown type in \textsc{IALam2}.

@@

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Another approach is to mark an error around the entire
application expression.

@@

Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
the marked conditional from the results.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{ISIf}

@@

However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
conditionals:
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
    \inconsistent{\TMV_1}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
  }{ISInconsistentBranches}

@@

@@

\subsection{Reified Hazelnut Action Semantics}
\label{sec:calculus-hazelnut}
